Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2019-08-01 ms/manuscript.Rmd
Update MS

```{r climate-numbers}
# Count total number of days in climate data
total_days <- daily_mean_climate %>% pull(date) %>% n_distinct()

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers. We calculated vapor pressure deficit (VPD) from relative humidity and temperature as implemented in the \package{plantecophys} R package [@Duursma2015].

@@

We used a multivariate approach to investigate differences in traits between epiphytes and terrestrial species. We used only quantitative traits, and included only species that had no missing data for any traits ($\n = 104$ \si{\species}). We first log-transformed traits as necessary to attain normality, then scaled each trait by its range. We subjected the transformed, scaled traits to a Principal Components Analysis (PCA) to reduce dimensionality. We conducted standard and phylogenetically corrected PCA, using the \function{PCA} and \function{phyl.pca} functions in the \package{FactoMineR} [@Le2008] and \package{phytools} R packages, respectively.

@@

We analyzed community functional trait diversity using the \function{dbFD} function in the \package{FD} R package [@Laliberte2011]. This function takes as input values any number of traits for species in a given community, plots them in a multidimensional trait space using PCA, and calculates several summary statistics. Functional richness (FRic) is equivalent to convex hull volume containing all observed trait values plotted in multidimensional trait space [@Cornwell2006], functional evenness (FEve) is the average distance to the center of all the traits, and functional dispersion (FDis) is the mean distance between traits [@Villeger2008]. We chose these three metrics because they are statistically independent and can each provide different information about the structure of community trait diversity [@Savage2012]. Functional richness does not use abundance data, but functional evenness and divergence can be weighted by abundance, so we preformed analyses for these metrics with abundance weighting to correct for the effect of rare species with unusual morphologies. Selection, transformation, and scaling of traits was performed similarly as for the PCA, except that the method used to calculate distance [@Gower1971] can accommodate missing data, so we included some species with observations missing for some traits ($\n = 126$ \si{\species} total).

@@

<!-- CHECK PCA CODE: missing dissection -->

@@

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "ntaxa", "p")`; \fig \@ref(fig:comm-div)). Species richness increased  with elevation for both epiphytes and terrestrial communities (\fig \@ref(fig:comm-div)). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "FRic", "p")`), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "FDiv", "p")`), and functional evenness (FEve) did not differ between the two. Both epiphytic and terrestrial communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "FRic", "r")`, \pval = `r get_mod_result(model_summ_print, "FRic", "p")`). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, \pval = `get_mod_result(t_test_print, "mntd.obs.z", "p")`) but not MPD (\pval = `r get_mod_result(t_test_print, "mpd.obs.z", "p")`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)).

>>>>

```{r climate-numbers, results="hide"}
# Count total number of days in climate data
total_days <- daily_mean_climate %>% pull(date) %>% n_distinct()

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers. We calculated vapor pressure deficit (VPD) from relative humidity and temperature using the \package{plantecophys} R package [@Duursma2015].

@@

```{r check-num-traits}
# Check the number of species included in trait PCA
cont_traits = c("sla", "stipe", "length", "width", 
                "rhizome", "pinna", "dissection")

quant_traits_nsp <-
fern_traits %>%
  # Subset to continuous traits
  select(species, habit, cont_traits) %>%
  # Keep only completely sampled species
  filter(complete.cases(.)) %>%
  nrow
```

We used a multivariate approach to investigate differences in traits between epiphytes and terrestrial species. We used only quantitative traits, and included only species that had no missing data for any traits (*\n* = `r quant_traits_nsp` \si{\species}). We first log-transformed traits as necessary to attain normality, then scaled each trait by its range. We subjected the transformed, scaled traits to a Principal Components Analysis (PCA) to reduce dimensionality. We conducted standard and phylogenetically corrected PCA, using the \function{PCA} and \function{phyl.pca} functions in the \package{FactoMineR} [@Le2008] and \package{phytools} R packages, respectively.

@@

We analyzed community functional trait diversity using the \function{dbFD} function in the \package{FD} R package [@Laliberte2011]. This function takes as input values any number of traits for species in a given community, plots them in a multidimensional trait space using PCA, and calculates several summary statistics. Functional richness (FRic) is equivalent to convex hull volume containing all observed trait values plotted in multidimensional trait space [@Cornwell2006], functional evenness (FEve) is the average distance to the center of all the traits, and functional dispersion (FDis) is the mean distance between traits [@Villeger2008]. We chose these three metrics because they are statistically independent and can each provide different information about the structure of community trait diversity [@Savage2012]. Functional richness does not use abundance data, but functional evenness and divergence can be weighted by abundance, so we preformed analyses for these metrics with abundance weighting to correct for the effect of rare species with unusual morphologies. Selection, transformation, and scaling of traits was performed similarly as for the PCA, except that the method used to calculate distance [@Gower1971] can accommodate missing data, so we included species with observations missing for some traits.

@@

@@

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "ntaxa", "p")`; \fig \@ref(fig:comm-div)). Species richness increased  with elevation for both epiphytes and terrestrial communities (\fig \@ref(fig:comm-div)). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "FRic", "p")`), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "FDiv", "p")`), and functional evenness (FEve) did not differ between the two. Both epiphytic and terrestrial communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, \rval^2^ = `r get_mod_result(model_summ_print, "FRic", "r")`, \pval = `r get_mod_result(model_summ_print, "FRic", "p")`). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, \pval = `r get_mod_result(t_test_print, "mntd.obs.z", "p")`) but not MPD (\pval = `r get_mod_result(t_test_print, "mpd.obs.z", "p")`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with elevation in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)).
